En garde! Unguarded iteration for reversible computation in the Delay monad\thanks{Niccol\`o Veltri was supported by a research grant (13156) from VILLUM FONDEN.}

### 2019-04-30 mpc19/paper.tex

An Elgot monad \cite{GoncharovMR16,GoncharovSRJ18} is a monad $T$
whose Kleilsi category supports unguarded uniform iteration. More
precisely\footnote{Here we give the definition of Elgot monad on
  $\Set$, but the definition of Elgot monad makes sense in any
  category with binary coproducts.}, a monad $\T$ is Elgot if there
exists an operation
\[
\iter_\T : (A \to \T \,(B + A)) \to A \to \T\,B
\]
satisfying a number of axioms:
\[\arraycolsep=8pt\def\arraystretch{1.8}
\begin{array}{c}
\mathsf{fixpoint} :
\iter_\T\,f \equiv \copair {\eta_\T}{\iter_\T \,f} \diamond_\T f
\\
\mathsf{naturality} : g \diamond_\T \iter_\T\,f \equiv
  \iter_\T\,(\copair{\T \inl \comp g} {\eta \comp \inr} \diamond_\T f)
\\
\mathsf{codiagonal} :
\iter_\T \,(\iter_\T\,g) \equiv \iter_\T\,(\T \copair {\id} {\inr}
  \comp g)
\\[.25cm]
\infer{\mathsf{uniformity}\,p : \iter_\T \,f \comp h \equiv \iter_\T
  \,g}{p : f \comp h \equiv \T (\id + h) \comp g}
\end{array}
\]
where $\eta_\T$ is the unit of $\T$ and $\diamond_\T$ denotes morphism
composition in the Kleisli category of $\T$.
The standard definition of uniform iteration operator includes the
dinaturality axiom, which has recently been discovered to be derivable
from the other laws \cite{EsikG16,GoncharovMR16}.

@@

Though the Kleisli category of the delay monad on $\Set$ is well studied,
comparatively less is known about this monad on other categories. It could be
interesting to study under which conditions its iterator exist -- e.g., whether
this is still the case when $\Set$ is replaced with an arbitrary topos. Another
avenue concerns the study of \emph{time invertible} programming languages:
Though not immediately clear in the current presentation, the trace on
$\Inv\Dapprox$ is not just reversible (in the sense that it preserves partial
isomorphisms) but in fact time invertible, in the sense that the number of
computation steps needed to perform $\trace (\dagger f)$ is \emph{precisely}
the same as what is needed to perform $\trace~f$ on any input. Since the
delay monad conveniently allows the counting of computation steps, we
conjecture that this is an ideal setting in which to study such intentional
semantics of reversible programming languages.

>>>>

An Elgot monad \cite{GoncharovMR16,GoncharovSRJ18} is a monad $T$
whose Kleilsi category supports unguarded uniform iteration. More
precisely\footnote{Here we give the definition of Elgot monad on
  $\Set$, but the definition of Elgot monad makes sense in any
  category with binary coproducts.}, a monad $\T$ is Elgot if there
exists an operation
\[
\iter_\T : (A \to \T \,(B + A)) \to A \to \T\,B
\]
satisfying the following axioms:
\[\arraycolsep=8pt\def\arraystretch{1.8}
\begin{array}{c}
\mathsf{fixpoint} :
\iter_\T\,f \equiv \copair {\eta_\T}{\iter_\T \,f} \diamond_\T f
\\
\mathsf{naturality} : g \diamond_\T \iter_\T\,f \equiv
  \iter_\T\,(\copair{\T \inl \comp g} {\eta \comp \inr} \diamond_\T f)
\\
\mathsf{codiagonal} :
\iter_\T \,(\iter_\T\,g) \equiv \iter_\T\,(\T \copair {\id} {\inr}
  \comp g)
\\[.25cm]
\infer{\mathsf{uniformity}\,p : \iter_\T \,f \comp h \equiv \iter_\T
  \,g}{p : f \comp h \equiv \T (\id + h) \comp g}
\end{array}
\]
where $\eta_\T$ is the unit of $\T$ and $\diamond_\T$ denotes morphism
composition in the Kleisli category of $\T$.
The standard definition of uniform iteration operator includes the
dinaturality axiom, which has recently been discovered to be derivable
from the other laws \cite{EsikG16,GoncharovMR16}.

@@

Though the Kleisli category of the delay monad on $\Set$ is well studied,
comparatively less is known about this monad on other categories. It could be
interesting to study under which conditions its iterator exist -- e.g., whether
this is still the case when $\Set$ is replaced with an arbitrary topos. Another
avenue concerns the study of \emph{time invertible} programming languages:
Though not immediately clear in the current presentation, the trace on
$\Inv\Dapprox$ is not just reversible (in the sense that it preserves partial
isomorphisms) but in fact time invertible, in the sense that the number of
computation steps needed to perform $\traceD \,(\daggerD\, f)$ is \emph{precisely}
the same as what is needed to perform $\traceD\,f$ on any input. Since the
delay monad conveniently allows the counting of computation steps, we
conjecture that this is an ideal setting in which to study such intentional
semantics of reversible programming languages.
